Temporal ordering and registration of cross-sectional imaging data

### 2014-05-13 image_analysis_paper/pnas_draft2.tex
edited draft

Our first data set, shown in Figure~\ref{subfig:raw_data1}, consists of $49$ images from stage 5 of {\em Drosophila} embryogenesis.
%
Each image is from a different embryo imaged in a different orientation and fixed at a different developmental time. 
%
We registered and ordered the images using vector diffusion maps.
%
The results are shown in Figure~\ref{subfig:ordered_data1}.
%
The features that our eyes can visually detect (the green Dorsal peak, and the red dpERK peaks) are now consistently registered, even though the algorithm does not require any information about the existence or location of these peaks. 
%
In addition, the red dpERK peaks grow in intensity over the developmental trajectory, as expected.

When dealing with large sets of images with interembryo variability, we can gain a better understanding of the developmental trajectory by looking at an averaged trajectory.
%
Figure~\ref{subfig:average_data1} shows the moving average of the trajectory in Figure~\ref{subfig:ordered_data1}. 
%
We can now easily see the developmental dynamics: the red dpERK peaks grow in time, and the ventral furrow (the dent in green) begins to form at the end of stage 5.

For this data set, we can validate our proposed vector diffusion maps ordering of the data.
%
During stage 5, we can measure the developmental time based on the monotonic progress of cellularization (details are given in the {\it SI Appendix}).
%
We can compare the temporal ordering obtained from cellularization with the ordering we extract from vector diffusion maps. 
%
The rank correlation coefficient between these two orderings is 0.9127, indicating that there is very good agreement and vector diffusion maps is uncovering the correct temporal ordering of the images.

@@

The images in Figure~\ref{fig:data1} are relatively simple, and the methods presented here are perhaps unnecessarily complex for these images. 
%
These data are effectively a single mode growing in time.
%
Therefore, principal component analysis \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
%
However, developmental dynamics are often significantly more complex, and nonlinear techniques such as diffusion maps/vector diffusion maps are required to extract meaningful structure in the data. 
%
Figure~\ref{subfig:raw_data2} shows a data set of images from stages 5--7 of {\it Drosophila} embryogenesis.
%
Only 22\% of the variability in the data is captured by the first principal component, and so we are forced to use more sophisticated techniques, such as vector diffusion maps, to temporally order the data.

Figure~\ref{subfig:ordered_data2} shows the images in Figure~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
%
Unlike the images in Figure~\ref{subfig:raw_data1}, 
for Figure~\ref{fig:data2}, we have no time marker with which to validate our orderings.
%
However, there are noticeable features in the ordering that we can see are consistent with previously known dynamics.
%
We can again construct a smoothed trajectory by looking at the moving average of the images, shown in Figure~\ref{sub fig:average_data2}.
%
We can easily see the formation of the ventral furrow and germ band elongation, which concludes with ventral cells shifted to the dorsal side of the embryo.
%
This assures us that we can obtained an ordering of the data which is developmentally meaningful.

@@

The task of image registration has been studied in a variety of contexts \cite{...}.
%
However, most of these applications first the definition and identification of good {\it features} within each image. 
%
For our biological applications, good features may not be known {\it a priori}.
%
We therefore use an algorithm which requires no features, but instead, uses the images directly.

The task of temporal ordering has also been studied, for data such as ...
%
Most of these applications order the data by solving a traveling salesman problem or constructing a minimum spanning tree on the data,
expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
%
We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
%
Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
%
They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
% 
Furthermore, techniques such as diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data.
% 
In this particular setup, we expect our data to be one-dimensional, as the only (main) source of variability is time.
%
However, other sources of variability would also be uncovered using diffusion maps, 
and tasks such as classification and comparison of different mutations can also be accomplished using the same techniques.

>>>>

Our first data set, shown in Figure~\ref{subfig:raw_data1}, consists of $49$ images from stage 5 of {\em Drosophila} embryogenesis.
%
Each image is from a different embryo imaged in a different orientation and fixed at a different developmental time. 
%
We registered and ordered the images using vector diffusion maps.
%
The results are shown in Figure~\ref{subfig:ordered_data1}.
%
The features that our eyes can visually detect (the green Dorsal peak, and the red dpERK peaks) are now consistently registered, even though the algorithm does not require any information about the existence or location of these peaks.

When dealing with large sets of images with interembryo variability, we can gain a better understanding of the developmental dynamics by looking at an averaged trajectory.
%
Figure~\ref{subfig:average_data1} shows the moving average of the trajectory in Figure~\ref{subfig:ordered_data1}. 
%
We can now easily see the developmental dynamics: the red dpERK peaks grow in time, and the ventral furrow (the dent in green) begins to form at the end of stage 5.

For this data set, we can validate our proposed vector diffusion maps ordering of the data.
%
During stage 5, we can measure the developmental time of each embryo based on the monotonic progress of cellularization (details are given in the {\it SI Appendix}).
%
We compared the temporal ordering obtained from cellularization with the ordering we extract from vector diffusion maps. 
%
The rank correlation coefficient between these two orderings is 0.9127, indicating that there is very good agreement and vector diffusion maps is uncovering the correct temporal ordering of the images.

@@

The images in Figure~\ref{fig:data1} are relatively simple, and the methods presented here are perhaps unnecessarily complex for this particular data set. 
%
Because these data are effectively a single mode growing in time,
principal component analysis \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data.
%
However, developmental dynamics are often significantly more complex, and nonlinear techniques such as diffusion maps/vector diffusion maps are required to extract meaningful structure in the data. 
%
Figure~\ref{subfig:raw_data2} shows a data set of images from stages 5--7 of {\it Drosophila} embryogenesis.
%
Only 22\% of the variability in the data is captured by the first principal component, and so we are forced to use more sophisticated techniques, such as vector diffusion maps, to temporally order the data.

Figure~\ref{subfig:ordered_data2} shows the images in Figure~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
%
Again, visually detectable features, such as the green Twist peaks, are consistently aligned. 
%
Unlike the images in Figure~\ref{subfig:raw_data1}, 
for Figure~\ref{fig:data2}, we have no time marker with which to validate our orderings.
%
However, there are noticeable features in the ordering that we can see are consistent with previously known dynamics.
%
We can again construct a smoothed trajectory by looking at the moving average of the images, shown in Figure~\ref{sub fig:average_data2}.
%
We can see the formation of the ventral furrow and germ band elongation, which concludes with ventral cells shifted to the dorsal side of the embryo, which is consistent with the known picture of {\it Drosophila} embryogeneisis. 
%
This assures us that we can obtained an ordering of the data which is developmentally meaningful.

@@

The task of image registration has been studied in a variety of contexts \cite{...}.
%
However, most of these applications rely on the definition and identification of good {\it features} within each image. 
%
For our biological applications, good features may not be known {\it a priori}, 
and we instead use an algorithm which requires no features, but instead, uses the images directly.

The task of temporal ordering has also been studied, for data such as ...
%
Most of these applications order the data by solving a traveling salesman problem or constructing a minimum spanning tree on the data,
expecting that these structures (the path of the traveling salesman, or the minimum spanning tree) characterize the majority of the structure/variability within the data and provide an accurate encoding of the dynamical progression.
%
We, instead, use diffusion maps to construct a one-dimensional parameterization of the data.
%
Diffusion maps are one of many nonlinear dimensionality reduction techniques that have been recently developed \cite{Belkin2003, tenenbaum2000global, Donoho2003, Roweis2000}.
%
They have been shown to be more robust to noise than other path-based algorithms, such as isomap \cite{balasubramanian2002isomap}, and so we suspect they may perform better than the ordering algorithms used in previous work.
% 
Furthermore, techniques such as diffusion maps do not constrain us to one-dimensional parameterizations/organizations of the data.
% 
In this particular setup, we expect our data to be one-dimensional, as the only (main) source of variability is time.
%
However, other sources of variability would also be uncovered using (vector) diffusion maps, 
and tasks such as classification of embryos by size or viewing angle, and comparison of different mutations can also be accomplished using the same techniques.
